---
title: "Getting Started with scimesh"
author: "Tim Schaefer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scimesh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is scimesh?

**scimesh** is a headless, GPU-free software renderer for 3D meshes.
It works anywhere R works — no X11, no OpenGL, no GPU required.  It is
designed as a drop-in replacement for the rendering part of
[fsbrain](https://github.com/dfsp-spirit/fsbrain) when rgl/OpenGL is
unavailable (e.g. on macOS with broken XQuartz, or on HPC clusters).

All rendering is done in modern C++ and returns in-memory RGBA images
that can be saved to PNG or composited into publication figures.

## Installation

```r
# Install from GitHub
remotes::install_github("dfsp-spirit/scimesh")
```

For FreeSurfer neuroimaging data, also install:
```r
remotes::install_github("dfsp-spirit/freesurferformats")
```

## Quick Start: Render a Cube

```{r, eval=FALSE}
library(scimesh)

sphere <- generate_sphere(c(0, 0, 0), radius = 1.2,
                          segments = 32, color = c(0.9, 0.3, 0.2, 1.0))
cam <- camera_auto(sphere, direction = c(1.2, 0.8, 1))

# Flat-shaded sphere
img <- render_mesh(sphere$vertices, sphere$triangles,
    colors = sphere$colors, camera = cam,
    options = render_options(
        lights = list(
            list(position = c(0.5, 1.0, 0.8), intensity = 1.5),
            list(position = c(-0.5, 0.2, 0.6), intensity = 0.5))))

write_png(img, "sphere.png")
```

For neuroimaging visualisation (FreeSurfer data, anatomical views, cortex
masking), see the **neuromesh** package which builds on scimesh:
```r
remotes::install_github("dfsp-spirit/neuromesh")
```

## Anatomical Views

Standard neuroimaging camera views (lateral, medial, dorsal, ventral,
anterior, posterior) and the `vis.subject.morph.native()` one-shot
visualisation function are provided by the **neuromesh** package.
See its vignette for examples.

## Colour Bars

Horizontal or vertical colour bars are produced in pure R (no X11) and
can be composed alongside rendered images:

```r
cbar <- colorbar_horizontal(viridis_colormap,
    n_colors = 256, width = 600, height = 80,
    ticks = c(0, 0.5, 1),
    tick_labels = c("min", "mid", "max"),
    title = "Value")
```

## Anti-Aliasing

scimesh supports ordered-grid supersampling (SSAA).  Pass `aa_samples = 2L`
for 2×2 SSAA (renders internally at double resolution, downsamples by
box averaging):

```r
render_options(aa_samples = 2L)
```

Values of 1 (off), 2, or 4 are supported.  Higher values give smoother
edges but use proportionally more memory and time.

## Semi-Transparent Overlays

scimesh supports proper alpha blending with depth-sorted back-to-front
rendering.  This is useful for visualising nested surfaces (white matter,
pial) or for glass-brain effects.

**R example:**
```{r, eval=FALSE}
library(scimesh)

white <- freesurferformats::read.fs.surface("sub-01/surf/lh.white")
pial  <- freesurferformats::read.fs.surface("sub-01/surf/lh.pial")

nv <- nrow(white$vertices)

white_mesh <- list(
    vertices  = white$vertices,
    triangles = white$faces,
    colors    = matrix(c(0.7, 0.7, 0.7, 1.0), nv, 4, byrow = TRUE))

pial_mesh <- list(
    vertices  = pial$vertices,
    triangles = pial$faces,
    colors    = matrix(c(0.9, 0.3, 0.2, 0.35), nv, 4, byrow = TRUE))

cam <- camera_auto(pial_mesh, direction = c(-1, 0, 0.2))
img <- render_scene(list(white_mesh, pial_mesh), cam,
    render_options(width = 1200, height = 900,
        backface_culling = FALSE,
        specular_color = c(0.4, 0.4, 0.4, 1),
        shininess = 64))
write_png(img, "transparency_lh.png")
```

**C++ example:** See `examples/cpp/transparency/` in the repository.

## Wireframe Mode

```r
render_options(wireframe = TRUE,
    wireframe_color = c(0, 0, 0, 1))  # black edges
```

Edges are computed via barycentric distance testing inside the triangle
rasterizer — no separate line primitives needed.  Edge thickness adapts
to triangle size so small triangles don't disappear.

## Specular Highlights

Add a glossy sheen to surfaces:

```r
render_options(
    specular_color = c(0.4, 0.4, 0.4, 1),  # white highlight
    shininess = 64)                          # tight spot
```

| `shininess` value | Look |
|-------------------|------|
| 8–16 | Soft plastic |
| 32–64 | Shiny surface |
| 128 | Glass-like tight spot |

## Projection Type

Perspective (default) or orthographic (parallel projection, matching
rgl's `view3d(fov = 0)` convention):

```r
render_options(projection = "orthographic")
```

## Using Diverging Colormaps

For signed data (Z-scores, cluster maps), use the built-in
blue–white–red diverging colormap:

```r
my_cmap <- diverging_colormap
```

Or bring your own colormap function:
```r
my_cmap <- function(n) grDevices::hcl.colors(n, palette = "Blue-Red 3")
```

## Compositing Multiple Images

`compose_layout()` arranges rendered images in a grid with optional
per-row/per-column cropping to eliminate wasted whitespace:

```r
img1 <- render_mesh(...)
img2 <- render_mesh(...)
result <- compose_layout(list(img1, img2), nrow = 1L, crop = TRUE)
write_png(result, "side_by_side.png")
```

## Interoperability with Other Packages

### rgl / tmesh3d

Convert an rgl `tmesh3d` object (or any list with `vb` and `it`
components) to scimesh format **without requiring the rgl package**:

```r
tmesh <- list(
    vb = rbind(c(-1, -1, 1, 1), c(-1, 1, 1, -1), c(0, 0, 0, 0), c(1, 1, 1, 1)),
    it = rbind(c(1L, 1L), c(2L, 3L), c(3L, 4L)))
mesh <- mesh_from_rgl(tmesh)
img  <- render_mesh(mesh$vertices, mesh$triangles)
```

### freesurferformats

The **neuromesh** package provides FreeSurfer-aware loading, cortex
masking, anatomical views, and colour bar integration on top of scimesh.
For lower-level access, use the `freesurferformats` package directly:

```r
surf  <- freesurferformats::read.fs.surface("lh.white")
sulc  <- freesurferformats::read.fs.morph("lh.sulc")
label <- freesurferformats::read.fs.label("lh.cortex.label")
```

### C++ via libfs

scimesh bundles [libfs](https://github.com/dfsp-spirit/libfs) — a
header-only C++ library for FreeSurfer file I/O.  C++ examples in
the repository demonstrate reading surfaces, curv data, labels, and
MGH volumes using `fs::read_surf()`, `fs::read_curv_data()`, etc.

### Python (future)

All core rendering functions are in C++ (`Mesh`, `Scene`, `Camera`,
`RenderOptions`, `Renderer`).  A Python binding (e.g., via pybind11)
would be straightforward as the C++ API is already designed for
language-agnostic use.

## Procedural Geometry

scimesh provides C++ and R functions for generating primitive geometry:

```r
cube    <- generate_cuboid(c(0, 0, 0), c(1, 1, 1), c(1, 0, 0, 1))
pyramid <- generate_pyramid(c(0, 0, 0), c(0, 2, 0), 1, c(0, 1, 0, 1))
tetra   <- generate_tetrahedron(c(0, 1, 0), c(1, -0.5, 0),
            c(-1, -0.5, 1), c(-1, -0.5, -1), c(0, 0, 1, 1))
torus   <- generate_torus(c(0, 0, 0), 1.5, 0.4, 32, 16, c(1, 0.5, 0, 1))
plane   <- generate_plane(c(0, 0, 0), c(0, 1, 0), 2, 1)
```

## Mesh Transforms

Translate, scale, rotate, or apply arbitrary 4×4 matrices to meshes
(all in C++ — no vertex data copied to R):

```r
mesh <- translate_mesh(mesh, c(5, 0, 0))
mesh <- scale_mesh(mesh, 2.0)
mesh <- rotate_mesh(mesh, pi / 4, c(0, 0, 1))
mesh <- transform_mesh(mesh, my_4x4_matrix)
```

## Mesh I/O

scimesh can read and write standard 3D mesh file formats.  All readers
return a mesh descriptor list with `vertices`, `triangles`, and
optionally `normals`, `uv`, or `colors`.

```r
# Read a Wavefront OBJ file (with optional UVs and normals)
mesh <- read_obj("model.obj")

# Read a Stanford PLY file (with optional vertex colors)
mesh <- read_ply("model.ply")

# Read an STL file (binary or ASCII)
mesh <- read_stl("model.stl")

# Write a mesh to STL
write_stl(mesh, "output.stl")               # binary (default)
write_stl(mesh, "output.stl", format = "ascii")
```

The OBJ reader supports multi-shape files and texture coordinates.
The PLY reader supports per-vertex RGB colors.  STL writes preserve
vertex normals when available.

## Face Colors

Per-face coloring lets you assign a single color to every vertex of a
triangle.  This is useful for parcellation overlays or material
assignment:

```r
mesh <- generate_cuboid(c(0, 0, 0), c(1, 1, 1))

# 12 triangles, each gets a color
fc <- matrix(c(1, 0, 0, 1), nrow = 12, ncol = 4, byrow = TRUE)
img <- render_mesh(mesh$vertices, mesh$triangles, face_colors = fc)
```

When `face_colors` is provided, it takes precedence over vertex
`colors`.

## Texture Mapping

Load a texture image and assign UV coordinates to render textured
meshes:

```r
# Requires the 'png' package
library(png)
tex <- readPNG("texture.png")

# UV coordinates (Nx2, values 0-1) must match vertex order
mesh <- read_obj("textured.obj")  # OBJ reader extracts UVs

img <- render_mesh(mesh$vertices, mesh$triangles,
    uv = mesh$uv, texture = tex)
```

Bilinear texture sampling is used for smooth results.

## Lower-Level Rendering

Beyond `render_mesh()` and `render_scene()`, scimesh provides
lower-level functions for special use cases:

### Raw Triangles (no index buffer)

Render geometry where positions and colors are flat arrays with 3
vertices per triangle — useful for dynamically generated geometry:

```r
positions <- matrix(c(0,0,0, 1,0,0, 0.5,1,0,
                      0,0,1, 1,0,1, 0.5,1,1), ncol = 3, byrow = TRUE)
colors <- matrix(c(1,0,0,1, 0,1,0,1, 0,0,1,1,
                   1,1,0,1, 0,1,1,1, 1,0,1,1), ncol = 4, byrow = TRUE)
cam <- camera_auto(positions)
img <- render_triangles(positions, colors, cam)
```

### Points

Render point cloud data with depth-tested circular markers:

```r
pts <- matrix(rnorm(300), ncol = 3)
cols <- matrix(c(1, 0.5, 0, 1), nrow = nrow(pts), ncol = 4, byrow = TRUE)
img <- render_points(pts, cols, radius = 4)
```

### Lines (as Cylinders)

Render line segments as thin cylinders:

```r
from <- matrix(c(0,0,0, 0,0,0), ncol = 3, byrow = TRUE)
to   <- matrix(c(1,0,0, 0,1,0), ncol = 3, byrow = TRUE)
cols <- matrix(c(1,0,0,1, 0,1,0,1), ncol = 4, byrow = TRUE)
img <- render_lines(from, to, radii = 0.05, colors = cols, cam)
```

## Camera Utilities

### Auto-Framing

`camera_auto()` computes a camera that fits any mesh or vertex set:

```r
verts <- matrix(rnorm(900), ncol = 3)
cam <- camera_auto(verts, direction = c(1, 1, 1), fov = 45)
```

It accepts either an Nx3 matrix or a mesh descriptor list.

### Manual Camera

For full control, use `camera()`:

```r
cam <- camera(eye = c(0, 0, 10), center = c(0, 0, 0),
              up = c(0, 1, 0), fov = 30)
```

## Mesh Utilities

### Bounding Box

Compute or visualise the axis-aligned bounding box:

```r
mesh <- generate_torus(c(0, 0, 0), 1.5, 0.4, 32, 16)
bbox <- mesh_bbox(mesh)
bbox$min  # c(xmin, ymin, zmin)
bbox$max  # c(xmax, ymax, zmax)

# Generate a wireframe bounding box mesh for rendering
bbox_mesh <- generate_bbox(mesh, color = c(0, 0, 0, 1), radius = 0.02)
```

### Axis Arrows

Generate coloured XYZ axis arrows (red X, green Y, blue Z):

```r
axes_mesh <- generate_axes(center = c(0, 0, 0), size = 2)
# Render together with your mesh:
img <- render_scene(list(mesh, axes_mesh), cam)
```

## Image Stacking

Quick helpers for combining images:

```r
# Vertical stack (one below another)
result <- stack_horizontal(img1, img2, img3)

# Vertical stack
result <- stack_vertical(img1, img2)

# With a colorbar
result <- stack_horizontal(img1, img2,
    colorbar = colorbar_horizontal(viridis_colormap))
```

## Running the R Examples

The repository includes runnable R example scripts in `examples/R/`:

| Example | What it shows |
|---------|--------------|
| `spot_cow/run.R` | Textured OBJ mesh with multi-light setup and SSAO |
| `transparency/run.R` | Semi-transparent pial overlay on white matter |
| `dragon/run.R` | Stanford Dragon with specular highlights and 4x AA |
| `whole_brain_sulc_single_image/run.R` | Whole-brain sulcal depth with cortex masking |

Run all examples at once:

```sh
cd examples/R
bash run_all_R.sh
```

For C++ examples, see `examples/cpp/` and
[`docs/CPP_GETTING_STARTED.md`](https://github.com/dfsp-spirit/scimesh/blob/main/docs/CPP_GETTING_STARTED.md).

## FAQ

**Why is there no interactive 3D window?**
scimesh is a headless (off-screen) renderer.  It produces static images.
For interactive rotation or real-time animation you still need rgl/OpenGL.

**How fast is it?**
A typical cortical surface (~300k triangles) at 1200×900 with 2× SSAA
renders in 1–3 seconds on a modern CPU.

**What about volume data?**
scimesh renders 3D surface meshes.  Volume slice visualisation
(e.g., `volvis.lb.with.surface()` in fsbrain) is done entirely in R/magick
without 3D rendering — it works independently of the renderer backend.

**Can I use it from Python?**
Not yet — but the C++ core is language-agnostic.  A pybind11 wrapper
would be straightforward.

**My images look pixelated/jagged — how do I fix this?**
Enable anti-aliasing: `render_options(aa_samples = 2L)`.

**How do I get a transparent background?**
```r
render_options(background_color = c(0, 0, 0, 0))
```
