---
title: "Getting Started with scimesh"
author: "Tim Schaefer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scimesh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is scimesh?

**scimesh** is a fast, headless, GPU-free software renderer for 3D triangle meshes
that produces **publication-quality images** for papers, slides, and presentations.
It works anywhere R works — no X11, no OpenGL, no GPU required.

Applications span any field that works with 3D surfaces:
- **Neuroimaging**: cortical surface visualisation (FreeSurfer data)
- **Structural biology**: molecule surfaces from PDB files
- **Computer graphics**: Stanford models, procedural geometry, textured meshes
- **Engineering and simulation**: mesh-based scientific visualisation of any kind

All rendering is done in modern C++17 and returns in-memory RGBA images
that can be saved to PNG or composed into multi-panel figures.

While scimesh can serve as a drop-in renderer backend when rgl/OpenGL is
unavailable (e.g., on macOS without XQuartz, on HPC clusters, in CI containers),
it is a **general-purpose visualisation tool** — not tied to any specific domain
or package.


## Installation

```r
# Install from GitHub
remotes::install_github("dfsp-spirit/scimesh")
```

For neuroimaging applications using FreeSurfer data, also install:

```r
remotes::install_github("dfsp-spirit/freesurferformats")
```

For the full viridis colormap family (magma, inferno, cividis, etc.), you may
optionally install `viridisLite`:

```r
install.packages("viridisLite")
```

But this is not required — scimesh ships built-in `viridis_colormap()` and
`diverging_colormap()` functions that use only base R.

## Quick Start

```{r, eval=FALSE}
library(scimesh)

sphere <- generate_sphere(c(0, 0, 0), radius = 1.2,
                          segments = 32, color = c(0.9, 0.3, 0.2, 1.0))
cam <- camera_auto(sphere, direction = c(1.2, 0.8, 1))

# Flat-shaded sphere
img <- render_mesh(sphere$vertices, sphere$triangles,
    colors = sphere$colors, camera = cam,
    options = render_options(
        lights = list(
            list(position = c(0.5, 1.0, 0.8), intensity = 1.5),
            list(position = c(-0.5, 0.2, 0.6), intensity = 0.5))))

write_png(img, "sphere.png")
```


## Rendering Features

### Render Options Reference

All rendering is controlled via `render_options()`. The following parameters
are available:

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `width`, `height` | integer | 800, 600 | Output image dimensions in pixels |
| `shading` | `"smooth"` / `"flat"` | `"smooth"` | Per-vertex (smooth) or per-face (flat) shading |
| `backface_culling` | logical | `TRUE` | Skip triangles facing away from camera |
| `background_color` | RGBA vector | `c(0,0,0,0)` | Transparent black by default |
| `default_color` | RGBA vector | `c(0.7,0.7,0.7,1)` | Fallback when no per-vertex colors |
| `invert_normals` | logical | `FALSE` | Flip face orientation |
| `wireframe` | logical | `FALSE` | Render edges only |
| `wireframe_color` | RGBA vector | `c(0,0,0,1)` | Edge color in wireframe mode |
| `projection` | `"perspective"` / `"orthographic"` | `"perspective"` | Camera projection type |
| `specular_color` | RGBA vector | `c(0,0,0,0)` | Specular highlight color (off by default) |
| `shininess` | numeric | 0 | Gloss tightness (8–128) |
| `ambient` | numeric | 0.3 | Ambient light level (0–1) |
| `contrast` | numeric | 1.0 | S-curve contrast multiplier |
| `lights` | list or `NULL` | `NULL` | Custom light list (auto-default if `NULL`) |
| `fog_enabled` | logical | `FALSE` | Enable depth fog |
| `fog_start` | numeric | 0 | Near-plane fog start |
| `fog_end` | numeric | 1 | Far-plane fog end |
| `fog_color` | RGBA vector | `c(0,0,0,0)` | Fog color |
| `aa_samples` | integer (1, 2, or 4) | 1 | Ordered-grid supersampling factor |
| `ssao_enabled` | logical | `FALSE` | Enable screen-space ambient occlusion |
| `ssao_radius` | numeric | 16 | SSAO sample radius in pixels |
| `ssao_intensity` | numeric | 0.8 | SSAO occlusion strength (0–1) |
| `threads` | integer | 0 | Number of CPU threads (0 = auto) |
| `clip_planes` | list or `NULL` | `NULL` | Custom clipping planes |

The following sections cover the most important options in detail.

### Lighting and Shading

scimesh uses a Blinn-Phong shading model with support for multiple light
sources, specular highlights, ambient control, and contrast adjustments.

#### Ambient Lighting

The `ambient` parameter (default 0.3) controls how much light reaches
surfaces that face away from the light source.  Lower values produce
deeper shadows and more contrast:

```r
render_options(ambient = 0.15)  # deeper shadows
```

#### Multi-Light Setups

Explicit lights give you full control over direction, colour, and
intensity:

```r
render_options(
    ambient = 0.2,
    contrast = 1.1,
    lights = list(
        list(position = c(0.5, 1.0, 0.8), color = c(1, 0.97, 0.9, 1),
             intensity = 1.5),
        list(position = c(-0.5, 0.2, 0.6), color = c(0.4, 0.5, 0.8, 1),
             intensity = 0.5)))
```

#### Specular Highlights

Add a glossy sheen to surfaces:

```r
render_options(
    specular_color = c(0.4, 0.4, 0.4, 1),  # white highlight
    shininess = 64)                          # tight spot
```

| `shininess` value | Look |
|-------------------|------|
| 8–16 | Soft plastic |
| 32–64 | Shiny surface |
| 128 | Glass-like tight spot |

#### Contrast Adjustment

Pass `contrast` to `render_options()` to apply an S-curve contrast
stretch after shading.  Values > 1.0 push darks toward black and
lights toward white, increasing perceived contrast.  The default 1.0
means no change:

```r
render_options(contrast = 1.1)  # subtle S-curve
```

The formula applied is `(value - 0.5) * contrast + 0.5`, clamped to
`[0, 1]`.  Typical values are 1.1--1.2 for a gentle boost, or up to
1.5 for a dramatic look.

You can also apply contrast as post-processing to an existing image:

```r
img <- render_mesh(mesh$vertices, mesh$triangles)
img <- image_apply_contrast(img, contrast = 1.1)
write_png(img, "output.png")
```

### Camera

#### Auto-Framing

`camera_auto()` computes a camera that fits any mesh or vertex set:

```r
verts <- matrix(rnorm(900), ncol = 3)
cam <- camera_auto(verts, direction = c(1, 1, 1), fov = 45)
```

It accepts either an Nx3 matrix or a mesh descriptor list.

#### Manual Camera

For full control, use `camera()`:

```r
cam <- camera(eye = c(0, 0, 10), center = c(0, 0, 0),
              up = c(0, 1, 0), fov = 30)
```

#### Orbiting the Camera

`camera_orbit()` rotates a camera's eye and up vector around its
center by a given angle about an axis.  This is useful for generating
turntable-style frame sequences:

```r
cam <- camera_auto(mesh, direction = c(1, 1, 1))

for (i in seq_len(8)) {
    cam_i <- camera_orbit(cam, angle_degrees = 360 / 8 * (i - 1))
    img   <- render_mesh(mesh$vertices, mesh$triangles, camera = cam_i,
                         options = render_options(width = 600, height = 400))
    write_png(img, sprintf("frame_%04d.png", i - 1))
}
```

For more complex trajectories, replace `camera_orbit()` with your own
function — it only needs to set `camera$eye` and `camera$up`.

The resulting PNG frames can be assembled into a video with ffmpeg
or into an animated GIF with your tool of choice.  See
`examples/R/video_frames_orbit/` for a runnable R example, and
`examples/cpp/brain_video/` for a C++ version that renders 48
full-brain frames.

### Projection Type

Perspective (default) or orthographic (parallel projection, matching
rgl's `view3d(fov = 0)` convention):

```r
render_options(projection = "orthographic")
```

### Anti-Aliasing

scimesh supports ordered-grid supersampling (SSAA).  Pass `aa_samples = 2L`
for 2x2 SSAA (renders internally at double resolution, downsamples by
box averaging):

```r
render_options(aa_samples = 2L)
```

Values of 1 (off), 2, or 4 are supported.  Higher values give smoother
edges but use proportionally more memory and time.

### Wireframe Mode

```r
render_options(wireframe = TRUE,
    wireframe_color = c(0, 0, 0, 1))  # black edges
```

Edges are computed via barycentric distance testing inside the triangle
rasterizer — no separate line primitives needed.  Edge thickness adapts
to triangle size so small triangles don't disappear.

### Screen-Space Ambient Occlusion (SSAO)

SSAO adds contact shadows in crevices and concavities, dramatically
improving depth perception and realism. It's a screen-space
post-processing effect — no extra geometry needed:

```r
render_options(
    ssao_enabled = TRUE,
    ssao_radius = 12,       # sample radius in pixels
    ssao_intensity = 0.5)   # occlusion strength (0–1)
```

Higher `ssao_radius` values sample a larger area (more expensive but
softer shadows). Higher `ssao_intensity` darkens occluded regions more.
Typical settings: radius 8–16, intensity 0.4–0.8.

SSAO is demonstrated in the `spot_cow` and `dragon` R examples, and
extensively in the C++ examples.

### Depth Fog

Atmospheric depth fog fades distant geometry toward a background color,
useful for emphasising foreground objects or creating stylised renders:

```r
render_options(
    fog_enabled = TRUE,
    fog_start = 0.6,              # begin fade at 60 % depth
    fog_end = 1.0,                # fully fogged at far plane
    fog_color = c(0.9, 0.95, 1, 1))  # pale blue fog
```

Fog is applied linearly between `fog_start` and `fog_end` in
normalised device coordinates.

### Semi-Transparent Overlays

scimesh supports proper alpha blending with depth-sorted back-to-front
rendering.  This is useful for visualising nested surfaces (e.g., white
matter and pial surface in neuroimaging), or for glass-brain effects.

**R example:**
```{r, eval=FALSE}
library(scimesh)

white <- freesurferformats::read.fs.surface("sub-01/surf/lh.white")
pial  <- freesurferformats::read.fs.surface("sub-01/surf/lh.pial")

nv <- nrow(white$vertices)

white_mesh <- list(
    vertices  = white$vertices,
    triangles = white$faces,
    colors    = matrix(c(0.7, 0.7, 0.7, 1.0), nv, 4, byrow = TRUE))

pial_mesh <- list(
    vertices  = pial$vertices,
    triangles = pial$faces,
    colors    = matrix(c(0.9, 0.3, 0.2, 0.35), nv, 4, byrow = TRUE))

cam <- camera_auto(pial_mesh, direction = c(-1, 0, 0.2))
img <- render_scene(list(white_mesh, pial_mesh), cam,
    render_options(width = 1200, height = 900,
        backface_culling = FALSE,
        specular_color = c(0.4, 0.4, 0.4, 1),
        shininess = 64))
write_png(img, "transparency_lh.png")
```

**C++ example:** See `examples/cpp/transparency/` in the repository.

### Background and Transparency

By default, scimesh renders with a transparent background
(`background_color = c(0, 0, 0, 0)`), which is ideal for compositing.

For a solid background (e.g., white for papers), set:

```r
render_options(background_color = c(1, 1, 1, 1))  # solid white
```

The background color is written to the output image's alpha channel,
so transparent backgrounds survive PNG export and can be further
composited in tools like ImageMagick or layout packages.


## Working with Meshes

### Procedural Geometry

scimesh provides C++ and R functions for generating primitive geometry:

```r
cube    <- generate_cuboid(c(0, 0, 0), c(1, 1, 1), c(1, 0, 0, 1))
sphere  <- generate_sphere(c(0, 0, 0), radius = 1.2,
                           segments = 32, color = c(0.9, 0.3, 0.2, 1))
cyl     <- generate_cylinder(c(0, -1, 0), c(0, 1, 0), 0.5, 32,
                             c(0.1, 0.7, 0.3, 1))
cone    <- generate_cone(c(0, -1.2, 0), c(0, 1.2, 0), 0.6, 32,
                         c(0.9, 0.7, 0.1, 1))
pyramid <- generate_pyramid(c(0, 0, 0), c(0, 1.5, 0), 1,
                            c(0.7, 0.2, 0.8, 1))
tetra   <- generate_tetrahedron(c(-1, -0.5, -1), c(1, -0.5, -1),
            c(0, -0.5, 1), c(0, 1.2, 0), c(0.2, 0.8, 0.8, 1))
torus   <- generate_torus(c(0, 0, 0), 1.0, 0.35, 24, 12,
                          c(0.6, 0.4, 0.2, 1))
plane   <- generate_plane(c(0, 0, 0), c(0, 1, 0), 1.2, 0.8,
                          c(0.5, 0.5, 0.5, 1))
```

See `examples/R/primitives/run.R` for a gallery script that renders all
primitives side-by-side in both shaded and wireframe mode.

### Mesh I/O

scimesh can read and write standard 3D mesh file formats.  All readers
return a mesh descriptor list with `vertices`, `triangles`, and
optionally `normals`, `uv`, or `colors`.

```r
# Read a Wavefront OBJ file (with optional UVs and normals)
mesh <- read_obj("model.obj")

# Read a Stanford PLY file (with optional vertex colors)
mesh <- read_ply("model.ply")

# Read an STL file (binary or ASCII)
mesh <- read_stl("model.stl")

# Write a mesh to STL
write_stl(mesh, "output.stl")               # binary (default)
write_stl(mesh, "output.stl", format = "ascii")
```

The OBJ reader supports multi-shape files and texture coordinates.
The PLY reader supports per-vertex RGB colors.  STL writes preserve
vertex normals when available.

### Mesh Transforms

Translate, scale, rotate, or apply arbitrary 4x4 matrices to meshes:

```r
mesh <- translate_mesh(mesh, c(5, 0, 0))
mesh <- scale_mesh(mesh, 2.0)
mesh <- rotate_mesh(mesh, pi / 4, c(0, 0, 1))
mesh <- transform_mesh(mesh, my_4x4_matrix)
```

### Face Colors

Per-face coloring lets you assign a single color to every vertex of a
triangle.  This is useful for parcellation overlays or material
assignment:

```r
mesh <- generate_cuboid(c(0, 0, 0), c(1, 1, 1))

# 12 triangles, each gets a color
fc <- matrix(c(1, 0, 0, 1), nrow = 12, ncol = 4, byrow = TRUE)
img <- render_mesh(mesh$vertices, mesh$triangles, face_colors = fc)
```

When `face_colors` is provided, it takes precedence over vertex
`colors`.

### Texture Mapping

Load a texture image and assign UV coordinates to render textured
meshes:

```r
# Requires the 'png' package
library(png)
tex <- readPNG("texture.png")

# UV coordinates (Nx2, values 0-1) must match vertex order
mesh <- read_obj("textured.obj")  # OBJ reader extracts UVs

img <- render_mesh(mesh$vertices, mesh$triangles,
    uv = mesh$uv, texture = tex)
```

Bilinear texture sampling is used for smooth results.

### Mesh Utilities

#### Bounding Box

Compute or visualise the axis-aligned bounding box:

```r
mesh <- generate_torus(c(0, 0, 0), 1.5, 0.4, 32, 16)
bbox <- mesh_bbox(mesh)
bbox$min  # c(xmin, ymin, zmin)
bbox$max  # c(xmax, ymax, zmax)

# Generate a wireframe bounding box mesh for rendering
bbox_mesh <- generate_bbox(mesh, color = c(0, 0, 0, 1), radius = 0.02)
```

#### Axis Arrows

Generate coloured XYZ axis arrows (red X, green Y, blue Z):

```r
axes_mesh <- generate_axes(center = c(0, 0, 0), size = 2)
# Render together with your mesh:
img <- render_scene(list(mesh, axes_mesh), cam)
```


## Lower-Level Rendering

Beyond `render_mesh()` and `render_scene()`, scimesh provides
lower-level functions for special use cases.

### Raw Triangles (no index buffer)

Render geometry where positions and colors are flat arrays with 3
vertices per triangle — useful for dynamically generated geometry:

```r
positions <- matrix(c(0,0,0, 1,0,0, 0.5,1,0,
                      0,0,1, 1,0,1, 0.5,1,1), ncol = 3, byrow = TRUE)
colors <- matrix(c(1,0,0,1, 0,1,0,1, 0,0,1,1,
                   1,1,0,1, 0,1,1,1, 1,0,1,1), ncol = 4, byrow = TRUE)
cam <- camera_auto(positions)
img <- render_triangles(positions, colors, cam)
```

### Points

Render point cloud data with depth-tested circular markers:

```r
pts <- matrix(rnorm(300), ncol = 3)
cols <- matrix(c(1, 0.5, 0, 1), nrow = nrow(pts), ncol = 4, byrow = TRUE)
img <- render_points(pts, cols, radius = 4)
```

### Lines (as Cylinders)

Render line segments as thin cylinders:

```r
from <- matrix(c(0,0,0, 0,0,0), ncol = 3, byrow = TRUE)
to   <- matrix(c(1,0,0, 0,1,0), ncol = 3, byrow = TRUE)
cols <- matrix(c(1,0,0,1, 0,1,0,1), ncol = 4, byrow = TRUE)
img <- render_lines(from, to, radii = 0.05, colors = cols, cam)
```


## Image Utilities

### Colour Bars

scimesh can produce horizontal or vertical colour bars in pure R
(no X11), ready to be composed alongside rendered images:

```r
cbar <- colorbar_horizontal(viridis_colormap,
    n_colors = 256, width = 600, height = 80,
    ticks = c(0, 0.5, 1),
    tick_labels = c("min", "mid", "max"),
    title = "Value")
```

Both `colorbar_horizontal()` and `colorbar_vertical()` accept any
colormap, specified either as:
- **A function** returning hex colors (e.g., `viridis_colormap`,
  `diverging_colormap`, `grDevices::hcl.colors`, or
  `viridisLite::viridis`)
- **A vector of color strings** (e.g., `c("red", "white", "blue")`)

Built-in colormaps (no extra packages required):

```r
# Viridis (perceptually uniform, colourblind-friendly)
viridis_colormap(256)

# Blue-white-red diverging (for signed data like Z-scores)
diverging_colormap(256)

# Any base R palette via wrapper functions
my_cmap <- function(n) grDevices::hcl.colors(n, palette = "inferno")
cbar <- colorbar_horizontal(my_cmap)

# Or pass colors directly
cbar <- colorbar_horizontal(c("darkblue", "cyan", "yellow", "red"))
```

If you have `viridisLite` installed, you can use its full colormap
family (magma, inferno, plasma, cividis) directly:

```r
library(viridisLite)
cbar <- colorbar_horizontal(viridis)  # viridisLite function
```

See `examples/R/colormaps/run.R` for a complete demonstration.

### Compositing Multiple Images

`compose_layout()` arranges rendered images in a grid with optional
per-row/per-column cropping to eliminate wasted whitespace:

```r
img1 <- render_mesh(...)
img2 <- render_mesh(...)
result <- compose_layout(list(img1, img2), nrow = 1L, crop = TRUE)
write_png(result, "side_by_side.png")
```

### Image Stacking

Quick helpers for combining images:

```r
# Horizontal stack (side by side)
result <- stack_horizontal(img1, img2, img3)

# Vertical stack (one below another)
result <- stack_vertical(img1, img2)

# With a colorbar
result <- stack_horizontal(img1, img2,
    colorbar = colorbar_horizontal(viridis_colormap))
```


## Interoperability with Other Packages

### rgl / tmesh3d

Convert an rgl `tmesh3d` object (or any list with `vb` and `it`
components) to scimesh format **without requiring the rgl package**:

```r
tmesh <- list(
    vb = rbind(c(-1, -1, 1, 1), c(-1, 1, 1, -1), c(0, 0, 0, 0), c(1, 1, 1, 1)),
    it = rbind(c(1L, 1L), c(2L, 3L), c(3L, 4L)))
mesh <- mesh_from_rgl(tmesh)
img  <- render_mesh(mesh$vertices, mesh$triangles)
```


## Examples

The repository includes runnable R example scripts in `examples/R/`:

| Example | What it shows |
|---------|--------------|
| `spot_cow/run.R` | Textured OBJ mesh with multi-light setup and SSAO |
| `dragon/run.R` | Stanford Dragon with specular highlights and 4x AA |
| `primitives/run.R` | All procedural primitives in shaded + wireframe gallery |
| `transparency/run.R` | Semi-transparent pial overlay on white matter |
| `whole_brain_sulc/run.R` | Whole-brain sulcal depth with cortex masking |
| `video_frames_orbit/run.R` | Turntable orbit frame sequence |
| `colormaps/run.R` | Custom colormaps with colorbars |

Run all examples at once:

```sh
cd examples/R
bash run_all_R.sh
```

For C++ examples, see `examples/cpp/` and
[`docs/CPP_GETTING_STARTED.md`](https://github.com/dfsp-spirit/scimesh/blob/main/docs/CPP_GETTING_STARTED.md).


## FAQ

**How fast is it?**
A typical cortical surface (~300k triangles) at 1200x900 with 2x SSAA
renders in 1--3 seconds on a modern CPU. Smaller meshes at lower
resolution can render in 200 ms or less, depending on the CPU and
number of lights.

**Why is there no interactive 3D window in which I can rotate the mesh?**
scimesh is a headless (off-screen) renderer, also known as a software
rasterizer. It produces images in roughly seconds, on just the CPU.
For interactive rotation or real-time animation, you need to render at
least 30 images *per second*. This is only achievable with a hardware
renderer, i.e., a graphics card and the full software stack required
to make use of it. If you have a graphics card, the typical solution
in R is to use rgl/OpenGL for interactive visualization.

**What about volume data, like 3D MRI scans in neuroimaging?**
scimesh renders 3D surface meshes. Volume slice visualisation
(e.g., `volvis.lb.with.surface()` in fsbrain) is done entirely in
R/magick without 3D rendering --- it works independently of the
renderer backend. Technically it's just a 2D image, no renderer
needed. This means you can use the functions in fsbrain for volume
visualisation even if you do not have rgl/OpenGL, and are using
scimesh for surface (mesh) rendering.

**My images look pixelated / jagged --- how do I fix this?**
Render at higher resolution and enable anti-aliasing:
`render_options(aa_samples = 2L)`.

**How do I get a transparent background?**
This is the default (`background_color = c(0, 0, 0, 0)`). See the
Background and Transparency section under Rendering Features.

**How do I control the number of CPU threads?**
Set the `threads` option in `render_options()`. The default 0
automatically uses all available cores. Set to 1 for single-threaded
rendering, or any positive integer to cap the thread count.

**What affects render performance?**
The main factors are triangle count, output resolution (pixels),
anti-aliasing level (`aa_samples`), number of lights, SSAO, and
whether transparency sorting is needed. For largest meshes, the most
effective optimisations are reducing `aa_samples` and output
resolution. SSAO and multi-light setups each add roughly constant
overhead per pixel.
