---
title: "Getting Started with scimesh"
author: "scimesh authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scimesh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is scimesh?

**scimesh** is a headless, GPU-free software renderer for 3D meshes.
It works anywhere R works — no X11, no OpenGL, no GPU required.  It is
designed as a drop-in replacement for the rendering part of
[fsbrain](https://github.com/dfsp-spirit/fsbrain) when rgl/OpenGL is
unavailable (e.g. on macOS with broken XQuartz, or on HPC clusters).

All rendering is done in modern C++ and returns in-memory RGBA images
that can be saved to PNG or composited into publication figures.

## Installation

```r
# Install from GitHub
remotes::install_github("dfsp-spirit/scimesh")
```

For FreeSurfer neuroimaging data, also install:
```r
remotes::install_github("dfsp-spirit/freesurferformats")
```

## Quick Start: Render a Brain

The simplest entry point is `vis.subject.morph.native()`, which loads
FreeSurfer surface data, maps per-vertex values to colors, and renders
standard anatomical views:

```{r, eval=FALSE}
library(scimesh)

img <- vis.subject.morph.native(
    subjects_dir = "~/data/study1",
    subject_id  = "subject1",
    measure     = "sulc",
    views       = "t4",
    cortex_only = TRUE,
    draw_colorbar = TRUE,
    colorbar_title = "Sulcal depth [mm]",
    aa_samples  = 2L                       # 2×2 supersampling anti-aliasing
)

write_png(img, "brain_sulc_t4.png")
```

This produces a tight 2×2 grid (lateral + medial, both hemispheres) with
a horizontal colour bar below.

## Anatomical Views

Pre-defined view shortcuts map directly to standard neuroimaging angles:

| Shortcut | Views included |
|----------|---------------|
| `"t4"` | lateral lh, lateral rh, medial lh, medial rh (2×2 grid) |
| `"t9"` | All 9 cardinal directions (3×3 grid) |
| `"si"` | Single interactive-equivalent (default lateral for given hemi) |

Single views are also available:
```r
vis.subject.morph.native(..., views = "lateral_lh")
vis.subject.morph.native(..., views = "medial_rh")
vis.subject.morph.native(..., views = "dorsal")
vis.subject.morph.native(..., views = "ventral")
vis.subject.morph.native(..., views = "anterior")
vis.subject.morph.native(..., views = "posterior")
```

## Colour Bars

Horizontal (default) or vertical colour bars are placed below or to the
right of the brain images.  Ticks and labels are auto-computed from the
data range using `pretty()`.

```r
# Horizontal colour bar (default)
vis.subject.morph.native(..., draw_colorbar = TRUE,
    colorbar_title = "Thickness [mm]")

# Vertical colour bar on the right
vis.subject.morph.native(..., draw_colorbar = "vertical",
    colorbar_title = "Thickness [mm]")

# Vertical colour bar on the left
# (via compose_layout's colorbar_side parameter)
```

## Anti-Aliasing

scimesh supports ordered-grid supersampling (SSAA).  Pass `aa_samples = 2L`
for 2×2 SSAA (renders internally at double resolution, downsamples by
box averaging):

```r
vis.subject.morph.native(..., aa_samples = 2L)
```

Values of 1 (off), 2, or 4 are supported.  Higher values give smoother
edges but use proportionally more memory and time.

## Semi-Transparent Overlays

scimesh supports proper alpha blending with depth-sorted back-to-front
rendering.  This is useful for visualising nested surfaces (white matter,
pial) or for glass-brain effects.

**R example:**
```{r, eval=FALSE}
library(scimesh)

white <- freesurferformats::read.fs.surface("sub-01/surf/lh.white")
pial  <- freesurferformats::read.fs.surface("sub-01/surf/lh.pial")

nv <- nrow(white$vertices)

white_mesh <- list(
    vertices  = white$vertices,
    triangles = white$faces,
    colors    = matrix(c(0.7, 0.7, 0.7, 1.0), nv, 4, byrow = TRUE))

pial_mesh <- list(
    vertices  = pial$vertices,
    triangles = pial$faces,
    colors    = matrix(c(0.9, 0.3, 0.2, 0.35), nv, 4, byrow = TRUE))

cam <- camera_auto(pial_mesh, direction = c(-1, 0, 0.2))
img <- render_scene(list(white_mesh, pial_mesh), cam,
    render_options(width = 1200, height = 900,
        backface_culling = FALSE,
        specular_color = c(0.4, 0.4, 0.4, 1),
        shininess = 64))
write_png(img, "transparency_lh.png")
```

**C++ example:** See `examples/cpp/transparency/` in the repository.

## Wireframe Mode

```r
render_options(wireframe = TRUE,
    wireframe_color = c(0, 0, 0, 1))  # black edges
```

Edges are computed via barycentric distance testing inside the triangle
rasterizer — no separate line primitives needed.  Edge thickness adapts
to triangle size so small triangles don't disappear.

## Specular Highlights

Add a glossy sheen to surfaces:

```r
render_options(
    specular_color = c(0.4, 0.4, 0.4, 1),  # white highlight
    shininess = 64)                          # tight spot
```

| `shininess` value | Look |
|-------------------|------|
| 8–16 | Soft plastic |
| 32–64 | Shiny surface |
| 128 | Glass-like tight spot |

## Projection Type

Perspective (default) or orthographic (parallel projection, matching
rgl's `view3d(fov = 0)` convention):

```r
render_options(projection = "orthographic")
```

## Using Diverging Colormaps

For signed data (Z-scores, cluster maps), use the built-in
blue–white–red diverging colormap:

```r
vis.subject.morph.native(...,
    makecmap_options = list(
        colFn = diverging_colormap,
        symm  = TRUE))
```

Or bring your own colormap function:
```r
my_cmap <- function(n) grDevices::hcl.colors(n, palette = "Blue-Red 3")
vis.subject.morph.native(..., makecmap_options = list(colFn = my_cmap))
```

## Compositing Multiple Images

`compose_layout()` arranges rendered images in a grid with optional
per-row/per-column cropping to eliminate wasted whitespace:

```r
img1 <- render_mesh(...)
img2 <- render_mesh(...)
result <- compose_layout(list(img1, img2), nrow = 1L, crop = TRUE)
write_png(result, "side_by_side.png")
```

## Interoperability with Other Packages

### rgl / tmesh3d

Convert an rgl `tmesh3d` object (or any list with `vb` and `it`
components) to scimesh format **without requiring the rgl package**:

```r
tmesh <- list(
    vb = rbind(c(-1, -1, 1, 1), c(-1, 1, 1, -1), c(0, 0, 0, 0), c(1, 1, 1, 1)),
    it = rbind(c(1L, 1L), c(2L, 3L), c(3L, 4L)))
mesh <- mesh_from_rgl(tmesh)
img  <- render_mesh(mesh$vertices, mesh$triangles)
```

### freesurferformats

`vis.subject.morph.native()` internally loads surface meshes, morphometry
data, and cortex labels via `freesurferformats`.  For custom pipelines,
use the library directly:

```r
surf  <- freesurferformats::read.fs.surface("lh.white")
sulc  <- freesurferformats::read.fs.morph("lh.sulc")
label <- freesurferformats::read.fs.label("lh.cortex.label")
```

### C++ via libfs

scimesh bundles [libfs](https://github.com/dfsp-spirit/libfs) — a
header-only C++ library for FreeSurfer file I/O.  C++ examples in
the repository demonstrate reading surfaces, curv data, labels, and
MGH volumes using `fs::read_surf()`, `fs::read_curv_data()`, etc.

### Python (future)

All core rendering functions are in C++ (`Mesh`, `Scene`, `Camera`,
`RenderOptions`, `Renderer`).  A Python binding (e.g., via pybind11)
would be straightforward as the C++ API is already designed for
language-agnostic use.

## Procedural Geometry

scimesh provides C++ and R functions for generating primitive geometry:

```r
cube    <- generate_cuboid(c(0, 0, 0), c(1, 1, 1), c(1, 0, 0, 1))
pyramid <- generate_pyramid(c(0, 0, 0), c(0, 2, 0), 1, c(0, 1, 0, 1))
tetra   <- generate_tetrahedron(c(0, 1, 0), c(1, -0.5, 0),
            c(-1, -0.5, 1), c(-1, -0.5, -1), c(0, 0, 1, 1))
torus   <- generate_torus(c(0, 0, 0), 1.5, 0.4, 32, 16, c(1, 0.5, 0, 1))
plane   <- generate_plane(c(0, 0, 0), c(0, 1, 0), 2, 1)
```

## Mesh Transforms

Translate, scale, rotate, or apply arbitrary 4×4 matrices to meshes
(all in C++ — no vertex data copied to R):

```r
mesh <- translate_mesh(mesh, c(5, 0, 0))
mesh <- scale_mesh(mesh, 2.0)
mesh <- rotate_mesh(mesh, pi / 4, c(0, 0, 1))
mesh <- transform_mesh(mesh, my_4x4_matrix)
```

## FAQ

**Why is there no interactive 3D window?**  
scimesh is a headless (off-screen) renderer.  It produces static images.
For interactive rotation or real-time animation you still need rgl/OpenGL.

**How fast is it?**  
A typical cortical surface (~300k triangles) at 1200×900 with 2× SSAA
renders in 1–3 seconds on a modern CPU.

**What about volume data?**  
scimesh renders 3D surface meshes.  Volume slice visualisation
(e.g., `volvis.lb.with.surface()` in fsbrain) is done entirely in R/magick
without 3D rendering — it works independently of the renderer backend.

**Can I use it from Python?**  
Not yet — but the C++ core is language-agnostic.  A pybind11 wrapper
would be straightforward.

**My images look pixelated/jagged — how do I fix this?**  
Enable anti-aliasing: `render_options(aa_samples = 2L)` or
`vis.subject.morph.native(..., aa_samples = 2L)`.

**How do I get a transparent background?**  
```r
render_options(background_color = c(0, 0, 0, 0))
```
